---
title: "Imbalanced integration guidelines tutorial - R"
output: pdf_document
date: '2023-04-24'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  tidy.opts = list(width.cutoff = 60),
  tidy = TRUE
)
```


# Imbalanced integration guidelines tutorial - R 

This tutorial is based on the [Imbalanced integration guidelines](https://www.biorxiv.org/content/10.1101/2022.10.06.511156v3.full) in the Iniquitate manuscript. This example explores integration in the case of a severely imbalanced dataset - namely dataset number 7 from the [Tran et al. scRNA-seq integration benchmarking paper](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-019-1850-9#availability-of-data-and-materials). Current scRNA-seq integration techniques perform very poorly on this dataset, as illustrated by the results from Tran et al. when benchmarking on this data. As such, we shouldn't expect a strong integration performance when it comes to conserving biological heterogeneity, but we'll try our best to tune this as we pretend we don't know the ground-truth labels. 

We'll begin by illustrating the properties of this dataset comprised of two batches of mouse retinal cells. 

## Dataset exploration 

Start by downloading the data - both batches:
```{r}
library(googledrive)

# Create data directory
if (!dir.exists("data")){
  dir.create("data")  
}

# Install files if not already present
if (!(file.exists("data/tran_et_al_batch_1.h5ad"))) {    
  # Remove authorization for Google Drive
  drive_deauth()
  drive_user()
  
  # Download the files
  file_id_1 <- as_id("17-5w1_Cfepf_AWsFPAjjGEKzKBhuB5Hr")
  file_id_2 <- as_id("1H-lfDjIKJmRSY-A7cNX-mIevWsn5rm5e")
  file_1 <- drive_get(file_id_1)
  file_2 <- drive_get(file_id_2)
  googledrive::drive_download(
    file_1, 
    path = "data/tran_et_al_batch_1.h5ad", 
    overwrite=TRUE
  )
  googledrive::drive_download(
    file_2,
    path = "data/tran_et_al_batch_2.h5ad",
    overwrite=TRUE
  )
}
```

Read in both files using the SeuratDisk library
```{r}
library(SeuratDisk)
# Convert the tran et al files to h5seurat format
Convert(
  "data/tran_et_al_batch_1.h5ad", 
  dest = "h5seurat", 
  overwrite = TRUE
)
Convert(
  "data/tran_et_al_batch_2.h5ad",
  dest = "h5seurat",
  overwrite = TRUE
)

# Load h5ad files for both batches
retina_1 <- SeuratDisk::LoadH5Seurat("data/tran_et_al_batch_1.h5seurat")
retina_2 <- SeuratDisk::LoadH5Seurat("data/tran_et_al_batch_2.h5seurat")
```

Examine the cell-type distribution for each batch
```{r, results = "markup", out.width = "100%", fig.align= "center"}
library(Seurat)
library(ggplot2)
library(ggthemes)

# Grab the metadata
meta_1 <- retina_1@meta.data
meta_2 <- retina_2@meta.data

# Print and plot the cell-type distribution for each batch
print(table(meta_1$celltype))
ggplot(data = meta_1, aes(celltype,fill=as.factor(celltype))) + 
  geom_histogram(stat="count") +
  labs(
    fill = "Cell-type",
    x = "Cell-type",
    y = "Count in batch 1"
  ) +
  theme_few() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(strip.text.x = element_text(size = 16)) +
  theme(plot.title = element_text(size = 14)) +
  theme(axis.text.x = element_text(size = 16)) +
  theme(axis.text.y = element_text(size = 16)) +
  theme(legend.title = element_text(size = 16)) +
  theme(legend.text = element_text(size = 16))

print(table(meta_2$celltype))
ggplot(data = meta_2, aes(celltype,fill=as.factor(celltype))) + 
  geom_histogram(stat="count") +
  labs(
    fill = "Cell-type",
    x = "Cell-type",
    y = "Count in batch 2"
  ) +
  theme_few() +
  theme(axis.title.x = element_text(size = 16)) +
  theme(axis.title.y = element_text(size = 16)) +
  theme(strip.text.x = element_text(size = 16)) +
  theme(plot.title = element_text(size = 14)) +
  theme(axis.text.x = element_text(size = 16)) +
  theme(axis.text.y = element_text(size = 16)) +
  theme(legend.title = element_text(size = 16)) +
  theme(legend.text = element_text(size = 16))
```
As we can see, there is an extreme imbalance in terms of the cell-types present in both batches, and this is also a special case of imbalance where quite a few cell-types (astrocytes, fibroblasts, ganglion, horizontal, microglia, pericytes, vascular endothelium) are only present in batch 2 but not batch one.

Now that we've seen the imbalance present in this dataset, we're going to take the role of a single-cell analysis practitioner that **does not know the underlying cell-type distribution of this dataset**. I.e. we don't know what cell-types are present and our goal is to learn a joint representation of these two batches of retinal cells that we've obtained from an experiment.  
